---
title: "workshop_code"
output: html_document
date: '2022-03-23'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## File conversion. Most are command-line tools

TriFusion: https://github.com/ODiogoSilva/TriFusion
plink: https://zzz.bwh.harvard.edu/plink/
vcftools: https://vcftools.github.io/man_latest.html
PGDspider: http://www.cmpg.unibe.ch/software/PGDSpider/
Phyluce: https://phyluce.readthedocs.io/en/latest/index.html

Custom code repos: 
https://github.com/edgardomortiz/vcf2phylip
https://github.com/mgharvey/misc_python
https://github.com/mgharvey/seqcap_pop/

Tons of great info here:
https://grunwaldlab.github.io/Population_Genetics_in_R/index.html
https://yulab-smu.top/treedata-book/chapter1.html


# load packages

```{r, echo=FALSE}

# data manipulation and organization packages
library(here)
library(dplyr)
library(stringr)
library(randomcoloR)

# population genetics packages
library(genepop)
library(PopGenome)
library(adegenet)
library(poppr)
library(pegas)
library(hierfstat)

# phylogenetics packages
library(ape)
library(caper)
library(phytools)
library(phangorn)
library(ggtree)


# define working folder path
here::i_am("workshop_code.R")
```


## DNA sequence data

Example data alignment: 
NEXUS
begin data;
dimensions ntax=6 nchar=520;
format datatype=dna missing=? gap=-;
matrix
cryptoleucus_139_LSU25431_0   ttaggacccctgctgtctgaaaagagatcaaagctcaggaggaggctgtaaaacagccgaatgcactgcacagaaacatg
cryptoleucus_139_LSU25431_1   ttaggacccctgctgtctgaaaagagatcaaagctcaggaggaggctgtaaaacagccgaatgcactgcacagaaacatg
cryptoleucus_141_LSU74103_0   ttaggacccctgctgtctgaaaagagatcaaagctcaggaggaggctgtaaaacagccgaatgcactgcacagaaacatg
cryptoleucus_141_LSU74103_1   ttaggacccctgctgtctgaaaagagatcaaagctcaggaggaggctgtaaaacagccgaatgcactgcacagaaacatg
cryptoleucus_142_LSU7285_0    ttaggacccctgctgtctgaaaagagatcaaagctcaggaggaggctgtaaaacagccgaatgcactgcacagaaacatg
cryptoleucus_142_LSU7285_1    ttaggacccctgctgtctgaaaagagatcaaagctcaggaggaggctgtaaaacagccgaatgcactgcacagaaacatg
;
end;


# PopGenome package
https://cran.r-project.org/web/packages/PopGenome/vignettes/Whole_genome_analyses_using_VCF_files.pdf

For descriptive statistics of population differentiation (Fst, Gst, Dxy)
Works with nucleotide alignment data
Folder should include separate alignment files for each locus
Fast! This example is for 2193 loci

First, let's load in our data

```{r, echo=FALSE}

folder_name <- "data/Epinecrophylla-phased-nexus-alignment"
GENOME.class <- readData(folder_name, format = 'nexus')

```


Next, we need to assign individuals to populations
First, let's assign each individual to its own population 
This will give an overall measure of population differentiation within the species

```{r}

sp <- get.individuals(GENOME.class)[[2]]
sp[1:6]

# data are phased, so each individual is represented by two sequences (each = 2)
GENOME.class <- set.populations(GENOME.class, unname(
  split(sp, rep(1:(length(sp)/2), each = 2))))

# Fst / Gst
GENOME.class <- F_ST.stats(GENOME.class)

# these results are stored as a massive per-locus matrix
# so I've saved it to a dataframe and taken the mean across all loci for each statistic
res <- as.data.frame(get.F_ST(GENOME.class))
res[1:5,]
colMeans(res, na.rm = TRUE)

```


# Dxy

Dxy is the degree of population differentiation
and is calculated as the amount of nucleotide diversity between populations, per site

```{r}

GENOME.class <- diversity.stats.between(GENOME.class)
dxy <- mean(GENOME.class@nuc.diversity.between / GENOME.class@n.sites)

dxy
```

# statistics per population

These data are for a species complex, so let's re-assign
individuals to a priori populations, perhaps subspecies

```{r}

pops <- split(sp, c("1", "1", "2", "2", "6", "6", "3", "3", "3", "3", 
                    "1", "1", "3", "3", "4", "4", "2", "2", "2", "2", 
                    "1", "1", "6", "6", "5", "5", "4", "4", "1", "1",
                    "2", "2", "3", "3", "2", "2", "4", "4", "3", "3", 
                    "4", "4"))
names(pops) <- c("amazonica", "haematonota", "pyrrhonota", "spodionota", "dentei", "fjeldsaai")
GENOME.class <- set.populations(GENOME.class, pops)

# We can check that this worked:
GENOME.class@populations

# Fst / Gst
GENOME.class <- F_ST.stats(GENOME.class)
res.pairwise <- as.data.frame(GENOME.class@Nei.G_ST.pairwise)
Fst.res.pairwise <- rowMeans(res.pairwise, na.rm = TRUE)
round(Fst.res.pairwise, 3)


```



We can also calculate simple diversity statistics for each population
say, nucleotide diversity

```{r}

GENOME.class <- diversity.stats(GENOME.class)

# this command lists five diversity stats, if you're interested
# the slot at the end pulls the data for a given population (e.g. 1,2,3)
res.pop.t <- get.diversity(GENOME.class)
res.pop <- apply(res.pop.t, 1, function (x) colMeans(as.data.frame(x)))
round(res.pop, 3)

```



# SNP data

PopGenome can also work with SNP data, stored in the variant call format (.vcf)

```{r}

GENOME.class.vcf <- readData("data/vcf_snps", format = 'VCF')

sp.vcf <- get.individuals(GENOME.class.vcf)[[1]]
pops.vcf <- split(sp.vcf, c("1", "1", "1", "1", 
                    "2", "2", 
                    "3", "3", "3", "3", 
                    "4", "4", "4", "4",
                    "5", "5", "5", "5", 
                    "6", "6", "6", "6", "6", "6",
                    "7", "7", "7", "7", "7", "7", "7", "7"
                    ))
names(pops.vcf) <- c("amazonica", "dentei", "fjeldsaai", "gutturalis", "haematonota", "pyrrhonota", "spodionota")

GENOME.class.vcf <- set.populations(GENOME.class.vcf, pops.vcf)

GENOME.class.vcf <- F_ST.stats(GENOME.class.vcf)
GENOME.class.vcf@Nei.G_ST.pairwise


```


We can also measure neutrality statistics, such as Tajima's D,
which is a measure of selection

```{r}

GENOME.class.vcf <- neutrality.stats(GENOME.class.vcf, FAST=TRUE)

# save all stats
neutrality <- get.neutrality(GENOME.class.vcf) 

# print just Tajima's D
TajimaD <- GENOME.class.vcf@Tajima.D
round(TajimaD, 3)

```


Waterson's Theta (effective population size)

```{r}

round(GENOME.class.vcf@theta_Watterson, 3)

```


# introgression 

We can also calculate archaic introgression between populations from SNP data
using Patterson's D statistic (ABBA/BABA tests). 
This requires at least three populations, plus an outgroup. 

Phylogeny 

  |-------- gutturalis
--|
  |  |----- pyrrhonota
  |--|
     |  |-- haematonota
     |--|
        |-- fjeldsaai


```{r}

pops.temp <- pops.vcf[c("pyrrhonota", "haematonota", "fjeldsaai")]

GENOME.class.vcf <- set.populations(GENOME.class.vcf, pops.temp)
GENOME.class.vcf <- set.outgroup(GENOME.class.vcf, "Epinecrophylla_gutturalis_e26_USNM587338.2")

GENOME.class.vcf <- introgression.stats(GENOME.class.vcf, do.D=TRUE)
GENOME.class.vcf <- introgression.stats(GENOME.class.vcf, do.df=TRUE)

paste0("Pattersons D: ", round(GENOME.class.vcf@D, 3)) 
paste0("Martin’s f statistic (fraction of the genome that is admixed): ", round(GENOME.class.vcf@f, 3))
paste0("and this fraction correcting for pairwise differences: ",  round(GENOME.class.vcf@df, 3))

```



Although this method does work, I recommend other software, such as Dsuite, for this purpose.


PopGenome can also work with whole genome data, and sliding-window analyses,
plus calculate many more statistics
More information in the manual and vignettes:
https://cran.r-project.org/web/packages/PopGenome/vignettes/Whole_genome_analyses_using_VCF_files.pdf
https://cran.r-project.org/web/packages/PopGenome/PopGenome.pdf




# Genepop package
good for isolation-by-distance calculations
genepop format from .vcf:
https://github.com/mgharvey/seqcap_pop/blob/master/bin/genepop_from_vcf.py

Genepop format:
Pop
8.56	-61.57 knipolegus_orenocensis_46_COP882, 0202 0101 0202 0101 0101 0101 0202 0101 0101 0101 0101 0101 0202 0101 0102 0202 0101 0101 0101 0101 0102 0101 
Pop
5.48	-67.62 knipolegus_orenocensis_47_COP629, 0000 0101 0202 0101 0101 0101 0202 0101 0101 0101 0101 0101 0202 0101 0101 0202 0101 0102 0101 0101 0101 0101


# Isolation-by-distance

```{r}

# define the input and output files
locinfile <- 'data/Knipolegus_orenocensis_SNPs_GenePop.IBD.txt'
outfile <- paste0('results/Knipolegus_orenocensis_GenePop.ISO.e')
ibd(locinfile,
    outputFile = outfile, 
    statistic = 'e')

res <- read.table(outfile, sep = '\t', header = FALSE)
res.e <- res %>% dplyr::filter(str_detect(V1, "a = "))
paste0("IBD results for intercept (a) and slope (b): ", res.e[1,])

```


# adegenet and poppr

https://github.com/thibautjombart/adegenet/wiki/Tutorials

Note: adegenet and Genepop accept the same data format, but require different file extension names
The files are identical except for .txt vs .gen
adegenet can also accept structure and genetix files. 

```{r}

locinfile <- 'data/Knipolegus_orenocensis_SNPs_GenePop.IBD.by_subspecies.gen'
knipolegus.genind <- read.genepop(locinfile)

# we can convert to other formats
knipolegus.genepop <- genind2genpop(knipolegus.genind)
inds.df <- genind2df(knipolegus.genind)
dim(inds.df)
inds.df[1:10, 1:5]

sums <- summary(knipolegus.genind)
sums$n.by.pop

# summary information, from poppr package
head(locus_table(knipolegus.genind))

```


We can do many of the same descriptive statistics in these packages! 

```{r}

# test for Hardy-Weinberg equilibrium (pegas package)
knipolegus.hwt <- hw.test(knipolegus.genind, B=0)
knipolegus.hwt[1:10,]
round(mean(knipolegus.hwt[,3], na.rm = TRUE), 4)

```

```{r}

# per-locus Weir-Cockerham Fst (from pegas)
fst.locus <- Fst(as.loci(knipolegus.genind))
fst.locus[1:10,]
round(mean(fst.locus[,2], na.rm = TRUE), 3)

```

```{r}

# overall Weir-Cockerham Fst (from hierfstat)
fst.overall <- wc(knipolegus.genind)
fst.overall

# pairwise Fst with Nei's estimator
matFst <- genet.dist(knipolegus.genind, method = "Nei87")
matFst

```


# Amount of missing data per locus in each population

```{r}

missing.pop <- info_table(knipolegus.genind)

#show first ten loci
missing.pop[,1:10]

```


# Missingness per individual

```{r}
missing.ind <- info_table(knipolegus.genind, type = "ploidy")
missing.ind[,1:10]

```

# removing missing data

```{r}

# removes loci
knipolegus.trimmed.loci <- knipolegus.genind %>% 
  missingno("loci", cutoff = 0.10)
knipolegus.trimmed.loci

# removes individuals
knipolegus.trimmed.inds <- knipolegus.genind %>% 
  missingno("geno", cutoff = 0.10)
knipolegus.trimmed.inds

# trim loci after removing problematic individual
knipolegus.trimmed.all <- knipolegus.trimmed.inds %>% 
  missingno("loci", cutoff = 0.10)
knipolegus.trimmed.all

```



# Phylogenetics
Read in trees and edit


```{r}

tree <- read.newick(file="data/trees/Epinecrophylla-exabayes-treePL-75p-intree.dated.tre")
summary.phylo(tree)
plot.phylo(tree)

```

Those tip labels are pretty excessively long
Let's edit them

```{r}
tree$tip.label[2]
tree$tip.label <- str_replace(tree$tip.label, "Epinecrophylla_", "")
tree$tip.label <- str_replace(tree$tip.label, "_1", "")
plot.phylo(tree)

```



A bit about tree structure

```{r}

plot.phylo(tree, show.tip.label = FALSE)
tiplabels()
nodelabels()
edgelabels()

```



If we want to just look at one part of the tree, we can extract subclades



```{r}

node <- getMRCA(tree, c("gutturalis_e20_AMNH11921", "dentei_e6_MZUSP80591"))
tree.sub <- extract.clade(tree, node)
plot.phylo(tree.sub)

tree.sub$tip.label
tree.sub <- keep.tip(tree.sub, tree.sub$tip.label[13:35])
tree.sub$tip.label <- gsub("amazonica_e5_LSUMZ75291", "haematonota_e5_LSUMZ75291", tree.sub$tip.label)
tree.sub$tip.label <- gsub("leucophthalma_leucophthalma_e44_LSUMZ5392", 
                       "spodionota_e44_LSUMZ5392", tree.sub$tip.label)
plot.phylo(tree.sub)

write.tree(tree.sub, file = "data/trees/Epinecrophylla-exabayes-treePL-75p-intree.dated.subsampled.tre")

```
# more plotting options

```{r}

plot.phylo(tree.sub, show.tip.label = FALSE, type = "unrooted")
plot.phylo(tree.sub, type = "fan", cex=0.4)

tree.sub <- drop.tip(tree.sub, "gutturalis_e20_AMNH11921")
plot.phylo(tree.sub)


rand.colors <- randomColor(count = length(tree.sub$tip.label), hue = c("random"))
names(rand.colors) <- tree.sub$tip.label
plot.phylo(tree.sub, show.tip.label = FALSE, type = "fan", cex=0.4)
tiplabels(pch=20, col=rand.colors, cex=4)

```


# Tree distances

Calculate the Robinson-Foulds distance between two trees
Here, I've loaded two trees estimated from the same samples but using two 
different methods, ExaBayes and RAxML

```{r}

tree.exabayes <- read.nexus(file="data/trees/Epinecrophylla-exabayes-75.tre")
tree.raxml <- read.tree(file="data/trees/RAxML_bipartitions.Epinecrophylla-75percent-final.tre")

plot.phylo(tree.exabayes, show.tip.label = FALSE, type = "unrooted")
plot.phylo(tree.raxml, show.tip.label = FALSE, type = "unrooted")

RF <- RF.dist(tree.exabayes, tree.raxml)
wRF <- wRF.dist(tree.exabayes, tree.raxml)

paste0("RF distance: ", RF, ", and weighted RF distance: ", round(wRF, 6)) 

```


We can also make these plots look pretty and save the output
treeio and ggtree are better for this 
First, we'll need to re-root the tree on a specific tip or node. Here we'll use a node:

```{r}

tree.plot <- treeio::read.newick("data/trees/RAxML_bipartitions.Epinecrophylla-75percent-final.tre")
tree.plot <- treeio::read.nexus("data/trees/Figure_3B_snap.Epinecrophylla_toClade_x2_priors.MCC.tre")

to.root <- getMRCA(tree.plot, 
                   c("Myrmorchilus_strigilatus_LSUMZ18722_1", "Neoctantes_niger_L2749_1"))
tree.plot <- root(tree.plot, node = to.root, resolve.root = TRUE)


# and plot
ggplot(tree.plot, aes(x, y, group=label)) + geom_tree() + theme_tree()
mapping = aes(
          x = branch,
          label = bootstrap
        )


structure <- read.csv("data/structure_data.csv", 
                      header = FALSE,
                      row.names = 1)


```


